Analysis-ready VCF at Biobank scale using Zarr

### 2024-05-30 paper.tex
Add some discussion of Cloud operation benefits

% We're not pretending is a solution to all problems and
% up front about weaknesses of the model.
Zarr provides pragmatic solutions to some of the more pressing 
problems facing the analysis of large-scale genetic variation
data, but it is not a panacea. Firstly, 
any dataset containing a variant with a large number of alleles
(perhaps due to indels) will cause problems because the 
dimensions of fields are determined by their \emph{maximum}
dimension among all variants. In particular this is problematic
for fields like PL in which the dimension depends 
quadratically on the number of alleles (although practical
solutions have been suggested that can be 
adapted~\cite{poterba2024scalable}).
Secondly, the design of 
VCF Zarr emphasises efficiency of analysis for a fixed 
dataset, and does not consider how samples (and the 
corresponding novel variants) should be added.
Thirdly, Zarr is not suitable for handling 
unstructured data often included in VCF INFO 
fields, and would likely be a poor choice
for representing variant 
annotations, for example.

Nonetheless, there are numerous datasets that exist today
what would benefit from cloud-native storage, 
inherent parallelism and granular storage.
[Mention benefits of fine-grained access control to 
different parts of the dataset using native storage access
control mechanism?]

The VCF Zarr specification and scalable vcf2zarr conversion utility
that we have provided here are the starting points of such 
an ecosystem, but much more would be needed to make a viable 
alternative for everyday bioinformatics workflows. However, a few key
pieces of infrastructure would provide a starting point, and would 
be sufficient to greatly improve researcher productivity on 
large, centrally managed datasets such as Genomics England.
Firstly, a \texttt{vcfzarrtools} command line utility which 
implements a subset of \texttt{bcftools} functionality (such as 
\texttt{view} and \texttt{query}) would immediately speed up 
limited region and field specific queries by orders of magnitude,
and would provide compatibility with existing workflows.
Developing such a utility would be a non-trivial but routine
engineering task.

>>>>

% We're not pretending is a solution to all problems and
% up front about weaknesses of the model.
Zarr provides pragmatic solutions to some of the more pressing 
problems facing the analysis of large-scale genetic variation
data, but it is not a panacea. Firstly, 
any dataset containing a variant with a large number of alleles
(perhaps due to indels) will cause problems because the 
dimensions of fields are determined by their \emph{maximum}
dimension among all variants. In particular this is problematic
for fields like PL in which the dimension depends 
quadratically on the number of alleles (although practical
solutions have been suggested that we plan to 
implement~\cite{poterba2024scalable}).
Secondly, the design of 
VCF Zarr emphasises efficiency of analysis for a fixed 
dataset, and does not consider how samples (and the 
corresponding novel variants) should be added.
Thirdly, Zarr is not suitable for handling 
the unstructured data often included in VCF INFO 
fields, and would likely be a poor choice
for representing variant 
annotations, for example.

Nonetheless, there are numerous datasets that exist today
that would likely reap significant benefits from being deployed 
in a cloud-native fashion using Zarr. Object 
stores typically allow for individual objects (chunks, in
Zarr) to be associated with ``tags'', which can then be 
used to associate storage class, user access control
and encryption keys on a chunk-by-chunk basis. 
Aside from the performance benefits
we have focused on here provided by Zarr, the ability
to (for example) use high-performance storage for commonly
used data such as the variant position and 
more cost-effective storage classes
for infrequently used bulk QC data should provide 
significant operational benefits.
Granular access controls would similarly allow non-identifiable
variant-level data to be shared relatively freely,
with genotype and other data more tightly controlled
as required.
Even finer granularity is possible if samples are grouped according
to access level within chunks (i.e, by padding partially filled 
chunks and using an appropriate sample mask). 
Delegating access control to the cloud provider in this way
and providing client applications direct access to
the data in an efficiently analysable form, 
largely negates the 
need for web API access~\cite{kelleher2019htsget} 
and cryptographic container formats~\citep{senf2021crypt4gh}
in this setting.

The VCF Zarr specification and scalable \texttt{vcf2zarr} conversion utility
provided here are a necessary starting point for such cloud-native
biobank repositories, but significant investment and development
would be needed 
to provide a viable alternative to standard bioinformatics workflows. 
A few key pieces of relatively straightforward infrastructure, 
however, would provide a starting point  and be
sufficient to greatly improve researcher productivity on 
large, centrally managed datasets such as Genomics England.
Firstly, a ``\texttt{vcztools}'' command line utility which 
implements a subset of \texttt{bcftools} functionality (such as 
\texttt{view} and \texttt{query}) on a VCF Zarr dataset 
would speed up some common queries by orders of magnitude,
and provide compatibility with existing workflows.
Developing such a utility would be a non-trivial but routine
engineering task.
